---
title: "Figure2"
author: "Galina Erikson"
date: "2026-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load required libraries
library(ggplot2)
library(dplyr)

# 1. Prepare the dataset
df <- data.frame(
  Sample = c(
    "HepG2_10_ulATACv2_rep1", "HepG2_10_ulATACv2_rep2",
    "HepG2_100_ulATACv2_rep1", "HepG2_100_ulATACv2_rep2",
    "HepG2_1000_ulATACv2_rep1", "HepG2_1000_ulATACv2_rep2",
    "HepG2_5000_ulATACv2_rep1", "HepG2_5000_ulATACv2_rep2",
    "HepG2_5000_ulATACv2_noMix"
  ),
  Peaks = c(85, 86, 5350, 6486, 16137, 18291, 32969, 40349, 53151),
  Condition = c("10", "10", "100", "100", "1000", "1000", "5000", "5000", "5000"),
  Replicate = c("Rep 1", "Rep 2", "Rep 1", "Rep 2", "Rep 1", "Rep 2", "Rep 1", "Rep 2", "noMix")
)

# 2. Enforce the X-axis order (10 to 5000)
df$Condition <- factor(df$Condition, levels = c("10", "100", "1000", "5000"))

# CRITICAL FIX: Explicitly order Replicate so 'noMix' is plotted last
df$Replicate <- factor(df$Replicate, levels = c("Rep 1", "Rep 2", "noMix"))

# 3. Create the grouped barplot
ggplot(df, aes(x = Condition, y = Peaks, fill = Replicate)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), color = "black", width = 0.7) +
  # Add value labels (position_dodge width must match geom_bar perfectly to align)
  geom_text(aes(label = scales::comma(Peaks)), 
            position = position_dodge(width = 0.8), 
            vjust = -0.5, 
            size = 3.5, 
            fontface = "bold") +
  theme_classic(base_size = 14) +
  scale_fill_viridis_d(option = "D", end = 0.85) + 
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "HepG2 ulATACseq Peak Counts",
    x = "Cell number",
    y = "Number of Narrow Peaks",
    fill = "Replicate"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", size = 0.5)
  )

```


#EL4

```{r pressure, echo=FALSE}
library(ggplot2)
library(dplyr)

# 1. Prepare the EL4 dataset
df <- data.frame(
  Sample = c(
    "EL4_10_ulATACv2_rep1", "EL4_10_ulATACv2_rep2",
    "EL4_100_ulATACv2_rep1", "EL4_100_ulATACv2_rep2",
    "EL4_1000_ulATACv2_rep1", "EL4_1000_ulATACv2_rep2",
    "EL4_5000_ulATACv2_rep1", "EL4_5000_ulATACv2_rep2",
    "EL4_5000_ulATACv2_noMix"
  ),
  Peaks = c(678, 246, 3845, 2286, 5064, 3829, 28912, 29355, 19814),
  Condition = c("10", "10", "100", "100", "1000", "1000", "5000", "5000", "5000"),
  Replicate = c("Rep 1", "Rep 2", "Rep 1", "Rep 2", "Rep 1", "Rep 2", "Rep 1", "Rep 2", "noMix")
)

# 2. Enforce the X-axis cell input order
df$Condition <- factor(df$Condition, levels = c("10", "100", "1000", "5000"))

# 3. Enforce the Replicate order so 'noMix' is plotted last
df$Replicate <- factor(df$Replicate, levels = c("Rep 1", "Rep 2", "noMix"))

# 4. Create the grouped barplot
ggplot(df, aes(x = Condition, y = Peaks, fill = Replicate)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), color = "black", width = 0.7) +
  # Add perfectly aligned text labels over the bars
  geom_text(aes(label = scales::comma(Peaks)), 
            position = position_dodge(width = 0.8), 
            vjust = -0.5, 
            size = 3.5, 
            fontface = "bold") +
  theme_classic(base_size = 14) +
  scale_fill_viridis_d(option = "D", end = 0.85) + 
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "EL4 ulATACseq Peak Counts",
    x = "Cell number",
    y = "Number of Narrow Peaks",
    fill = "Replicate"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", size = 0.5)
  )
```

#PCA
```{r pressure, echo=FALSE}
library(ggplot2)

pca <- read.table(
"/data/jenuwein/erikson/2026/ulATAC/Project_3925/m39_AOS/bw/Eingenvectors.tab",
header = TRUE,
row.names = 1
)

#Extract eigenvalues

eig <- as.numeric(pca$Eigenvalue)
var_explained <- eig/sum(eig) * 100

#Transpose and remove Eigenvalue row

pca_t <- as.data.frame(t(pca))
pca_t <- pca_t[rownames(pca_t) != "Eigenvalue", ]

#pca_t <- as.data.frame(t(pca[, 2:10])) # Exclude 'Component' and 'Eigenvalue' columns
colnames(pca_t) <- c("PC1", "PC2", "PC3") # Name the columns based on the rows

#2. Extract conditions from the new sample names (row names)

pca_t$condition <- case_when(
grepl("10", rownames(pca_t)) ~ "10",
grepl("100", rownames(pca_t)) ~ "100",
grepl("1000", rownames(pca_t)) ~ "1000",
grepl("5000", rownames(pca_t)) & !grepl("noMix", rownames(pca_t)) ~ "5000",
grepl("noMix", rownames(pca_t)) ~ "Mix",
TRUE ~ NA_character_
)

#Set the discrete factor levels for clean ordering in the legend

pca_t$condition <- factor(
pca_t$condition,
levels = c("10", "100", "1000", "5000", "Mix")
)

#3. Ensure PC coordinates are numeric

pca_t$PC1 <- as.numeric(pca_t$PC1)
pca_t$PC2 <- as.numeric(pca_t$PC2)

#4. Calculate variance explained using the 'Eigenvalue' column from your raw data
#(PC variance / total variance)

eig <- as.numeric(pca$Eigenvalue)
var_explained <- eig/sum(eig) * 100

#5. Plot with custom colors for each of the 5 groups

p <- ggplot(pca_t, aes(x = PC1, y = PC2, color = condition)) +
geom_point(size = 4, alpha = 0.8) +
scale_color_manual(values = c(
"10" = "#E41A1C", # Red
"100" = "#377EB8", # Blue
"1000" = "#4DAF4A", # Green
"5000" = "#984EA3", # Purple
"Mix" = "#FF7F00" # Orange
)) +
labs(
color = "Condition",
x = sprintf("PC1 (%.1f%%)", var_explained[1]),
y = sprintf("PC2 (%.1f%%)", var_explained[2])
) +
theme_classic()

print(p)

#1. Preferred Option: PDF (Vector format, perfect for publication text/plots)

ggsave(
filename = "Fig2.panelB.pca.pdf",
plot = p,
device = "pdf",
width = 6, # Width in inches
height = 5, # Height in inches
units = "in"
)

ggsave(
filename = "Fig2.panelB.pca.png",
plot = p,
device = "png",
width = 6,
height = 5,
units = "in",
dpi = 300 # High resolution requirement
)


```

#PCA
```{r pressure, echo=FALSE}
library(ggplot2)

# 1. Transpose the data so samples are rows and PCs are columns
# (Assuming your input dataframe is named 'pca_raw')

pca <- read.table(
  "/data/jenuwein/erikson/2026/ulATAC/Project_3925/m39_AOS/bw/Eingenvectors.tab",
  header = TRUE,
  row.names = 1
)

# Extract eigenvalues
eig <- as.numeric(pca$Eigenvalue)
var_explained <- eig/sum(eig) * 100

# Transpose and remove Eigenvalue row
pca_t <- as.data.frame(t(pca))
pca_t <- pca_t[rownames(pca_t) != "Eigenvalue", ]

#pca_t <- as.data.frame(t(pca[, 2:10])) # Exclude 'Component' and 'Eigenvalue' columns
colnames(pca_t) <- c("PC1", "PC2", "PC3")  # Name the columns based on the rows

# 2. Extract conditions from the new sample names (row names)
pca_t$condition <- case_when(
  grepl("_10_", rownames(pca_t)) ~ "10",
  grepl("_100_", rownames(pca_t)) ~ "100",
  grepl("_1000_", rownames(pca_t)) ~ "1000",
  grepl("_5000_", rownames(pca_t)) & !grepl("noMix", rownames(pca_t)) ~ "5000",
  grepl("noMix", rownames(pca_t)) ~ "Mix",
  TRUE ~ NA_character_
)

# Set the discrete factor levels for clean ordering in the legend
pca_t$condition <- factor(
  pca_t$condition,
  levels = c("10", "100", "1000", "5000", "Mix")
)

# 3. Ensure PC coordinates are numeric
pca_t$PC1 <- as.numeric(pca_t$PC1)
pca_t$PC2 <- as.numeric(pca_t$PC2)

# 4. Calculate variance explained using the 'Eigenvalue' column from your raw data
# (PC variance / total variance)

eig <- as.numeric(pca$Eigenvalue)
var_explained <- eig/sum(eig) * 100

p <- ggplot(pca_t, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 4, alpha = 0.8) +

  scale_color_manual(values = c(
    "10"    = "#E41A1C",
    "100"   = "#377EB8",
    "1000"  = "#4DAF4A",
    "5000"  = "#984EA3",
    "Mix"   = "#FF7F00"
  )) +

  labs(
    title = "EL4 ulATACseq PCA",
    color = "Condition",
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  ) +

  theme_classic(base_size = 14) +

  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 16
    ),
    axis.title = element_text(
      face = "bold"
    ),
    axis.text = element_text(
      size = 12
    ),
    legend.title = element_text(
      face = "bold"
    ),
    legend.text = element_text(
      size = 12
    ),
    panel.grid.major.y = element_line(
      color = "grey90",
      size = 0.5
    )
  )

print(p)
```

#Ploting the HepG2 data as well
```{r pressure, echo=FALSE}
library(ggplot2)

# 1. Transpose the data so samples are rows and PCs are columns
# (Assuming your input dataframe is named 'pca_raw')

pca <- read.table(
  "/data/jenuwein/erikson/2026/ulATAC/Project_3925/hg38_AOS/peakset/PCAeigenvalue.tab",
  header = TRUE,
  row.names = 1
)

samples_to_remove <- c(
  "HepG2_5000_ulATACv1_rep1.filtered.scalefac.bw", 
  "HepG2_5000_ulATACv1_rep2.filtered.scalefac.bw"
)
pca <- pca[, !(colnames(pca) %in% samples_to_remove)]

# Extract eigenvalues
eig <- as.numeric(pca$Eigenvalue)
var_explained <- eig/sum(eig) * 100

# Transpose and remove Eigenvalue row
pca_t <- as.data.frame(t(pca))
pca_t <- pca_t[rownames(pca_t) != "Eigenvalue", ]

#pca_t <- as.data.frame(t(pca[, 2:10])) # Exclude 'Component' and 'Eigenvalue' columns
colnames(pca_t) <- c("PC1", "PC2", "PC3")  # Name the columns based on the rows

# 2. Extract conditions from the new sample names (row names)
pca_t$condition <- case_when(
  grepl("_10_", rownames(pca_t)) ~ "10",
  grepl("_100_", rownames(pca_t)) ~ "100",
  grepl("_1000_", rownames(pca_t)) ~ "1000",
  grepl("_5000_", rownames(pca_t)) & !grepl("noMix", rownames(pca_t)) ~ "5000",
  grepl("noMix", rownames(pca_t)) ~ "Mix",
  TRUE ~ NA_character_
)

# Set the discrete factor levels for clean ordering in the legend
pca_t$condition <- factor(
  pca_t$condition,
  levels = c("10", "100", "1000", "5000", "Mix")
)

# 3. Ensure PC coordinates are numeric
pca_t$PC1 <- as.numeric(pca_t$PC1)
pca_t$PC2 <- as.numeric(pca_t$PC2)

# 4. Calculate variance explained using the 'Eigenvalue' column from your raw data
# (PC variance / total variance)

eig <- as.numeric(pca$Eigenvalue)
var_explained <- eig/sum(eig) * 100

p <- ggplot(pca_t, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 4, alpha = 0.8) +

  scale_color_manual(values = c(
    "10"    = "#E41A1C",
    "100"   = "#377EB8",
    "1000"  = "#4DAF4A",
    "5000"  = "#984EA3",
    "Mix"   = "#FF7F00"
  )) +

  labs(
    title = "HepG2 ulATACseq PCA",
    color = "Condition",
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2])
  ) +

  theme_classic(base_size = 14) +

  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 16
    ),
    axis.title = element_text(
      face = "bold"
    ),
    axis.text = element_text(
      size = 12
    ),
    legend.title = element_text(
      face = "bold"
    ),
    legend.text = element_text(
      size = 12
    ),
    panel.grid.major.y = element_line(
      color = "grey90",
      size = 0.5
    )
  )

print(p)
```

#Venn diagram
#erikson@maximus:/data/jenuwein/erikson/2026/ulATAC/ATAC_HepG2_Snyder_analysis/hg38_AOS/peakset$ bedtools intersect -a peaks.bed -b /data/jenuwein/erikson/2026/ulATAC/Project_3925/hg38_AOS/peakset/peaks.bed >common_peaks.bed


```{r pressure, echo=FALSE}
library(VennDiagram)

# Generate the fixed Venn Diagram
venn_plot <- draw.pairwise.venn(
  area1 = 187381,         
  area2 = 73852,          
  cross.area = 66055,     
  category = c("HEPG2 ENCODE\n(187,381)", "HEPG2 ulATAC\n(73,852)"),
  
  # --- THE FIXES FOR CUT-OFF TEXT ---
  margin = 0.1,                       # Adds 10% padding around the whole plot
  cat.pos = c(-15, 15),               # Adjusted angles to position text better
  cat.dist = c(0.08, 0.08),           # Distance of labels from circles
  # ----------------------------------
  
  # Esthetics and Styling
  fill = c("#440154FF", "#21908CFF"), 
  alpha = c(0.6, 0.6),                
  lty = "blank",                      
  
  # Text styling
  col = "white",                      
  cex = 1.5,                          
  cat.cex = 1.3                       
)

# Save it using the png() device approach (removes ggsave boundary limits)
png(
  filename = "Fig2.panelD.pca.png", 
  width = 6, 
  height = 5, 
  units = "in", 
  res = 300 
)

grid.draw(venn_plot)
dev.off()
```
#El4 peaks intersected with published data
#erikson@maximus:/data/jenuwein/erikson/2026/ulATAC/ATAC_EL4_public/aos/AOS/peakset$ bedtools intersect -a peaks.bed -b /data/jenuwein/erikson/2026/ulATAC/Project_3925_mergedReps_AOS/aos/peaks/EL4.unique.peaks >common_peaks.bed

```{r pressure, echo=FALSE}
library(VennDiagram)

# Generate the fixed Venn Diagram
venn_plot <- draw.pairwise.venn(
  area1 = 514625,         
  area2 = 74221,          
  cross.area = 60464,     
  category = c("EL4 GSE87646\n(514,625)", "EL4 ulATAC\n(74,221)"),
  
  # --- THE FIXES FOR CUT-OFF TEXT ---
  margin = 0.1,                       # Adds 10% padding around the whole plot
  cat.pos = c(-15, 15),               # Adjusted angles to position text better
  cat.dist = c(0.08, 0.08),           # Distance of labels from circles
  # ----------------------------------
  
  # Esthetics and Styling
  fill = c("#440154FF", "#21908CFF"), 
  alpha = c(0.6, 0.6),                
  lty = "blank",                      
  
  # Text styling
  col = "white",                      
  cex = 1.5,                          
  cat.cex = 1.3                       
)

# Save it using the png() device approach (removes ggsave boundary limits)
png(
  filename = "Fig2.panelD.EL4.venn.pca.png", 
  width = 6, 
  height = 5, 
  units = "in", 
  res = 300 
)

grid.draw(venn_plot)
dev.off()
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
