---
title: "PCA"
author: "Galina Erikson"
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading in data, make a PCA
Eingenvalues were calculated using: plotPCA --corData peakset/counts.bw.npz -o figures/PCA.png --transpose --ntop 5000
```{r}
pca <- read.table("/data/jenuwein/erikson/2026/ulATAC/bams_allData_withDiffATAC/bw/PCAeigenvalue.tab", header = TRUE, row.names = 1)
```

Plot
```{r}
pca_t <- as.data.frame(t(pca))
pca_t <- pca_t[rownames(pca_t) != "Eigenvalue", ]
pca_t$stage <- sub("_WT.*", "", rownames(pca_t))
pca_t$color <- with(pca_t, ifelse(stage == "Zygote", "red",
                          ifelse(stage == "X2Cells", "yellow",
                          ifelse(stage == "X8Cells", "green",
                          ifelse(stage == "Morula", "blue",
                          ifelse(stage == "Blastocyst", "purple", "black"))))))

library(ggplot2)

pca_t$PC1 <- pca_t[,1]
pca_t$PC2 <- pca_t[,2]
ggplot(pca_t, aes(x = PC1, y = PC2, color = stage)) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "Zygote" = "red",
    "X2Cells" = "yellow",
    "X8Cells" = "green",
    "Morula" = "blue",
    "Blastocyst" = "purple"
  )) +
  theme_classic()
```
#some modification
```{r}
pca_t$stage <- gsub("^X", "", pca_t$stage)
pca_t$stage <- factor(pca_t$stage,
                      levels = c("Zygote", "2Cells", "8Cells", "Morula", "Blastocyst"))
ggplot(pca_t, aes(x = PC1, y = PC2, color = stage)) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c(
      "Zygote" = "red",
      "2Cells" = "yellow",
      "8Cells" = "green",
      "Morula" = "blue",
      "Blastocyst" = "purple"
    ),
    breaks = c("Zygote", "2Cells", "8Cells", "Morula", "Blastocyst"),  # order
    labels = c("Zygote", "2Cell", "8Cell", "Morula", "Blastocyst")     # legend text
  ) +
  theme_classic()
```

#Adding the percentage to the plot
```{r}
library(ggplot2)

# 1. Read deepTools PCA matrix
pca <- read.table(
  "/data/jenuwein/erikson/2026/ulATAC/bams_allData_withDiffATAC/bw/PCAeigenvalue.tab", 
  header = TRUE, 
  row.names = 1,
  stringsAsFactors = FALSE
)

# 2. Extract Eigenvalues Safely
if ("Eigenvalue" %in% rownames(pca)) {
  # If Eigenvalue is a row
  eig_vals <- as.numeric(pca["Eigenvalue", ])
  pca_t <- as.data.frame(t(pca[rownames(pca) != "Eigenvalue", ]))
} else if ("Eigenvalue" %in% colnames(pca)) {
  # If Eigenvalue is a column
  eig_vals <- as.numeric(pca$Eigenvalue)
  pca_t <- as.data.frame(t(pca[, colnames(pca) != "Eigenvalue"]))
} else {
  # Fallback: assuming first row or column holds eigenvalues
  eig_vals <- as.numeric(pca[1, ])
  pca_t <- as.data.frame(t(pca[-1, ]))
}

# Remove any lingering NA values from eigenvalues
eig_vals <- eig_vals[!is.na(eig_vals)]

# Calculate variance percentage
var_explained <- (eig_vals / sum(eig_vals)) * 100

# 3. Format PC columns as numeric
pca_t$PC1 <- as.numeric(pca_t[, 1])
pca_t$PC2 <- as.numeric(pca_t[, 2])

# 4. Stage name formatting & ordering
pca_t$stage <- sub("_WT.*", "", rownames(pca_t))
pca_t$stage <- gsub("^X", "", pca_t$stage)

pca_t$stage <- factor(
  pca_t$stage,
  levels = c("Zygote", "2Cells", "8Cells", "Morula", "Blastocyst")
)

# 5. Plot
ggplot(pca_t, aes(x = PC1, y = PC2, color = stage)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(
    values = c(
      "Zygote"     = "#E41A1C", # Vivid Red
      "2Cells"     = "#FFD700", # Gold/Yellow (much more visible than pure yellow)
      "8Cells"     = "#4DAF4A", # Green
      "Morula"     = "#377EB8", # Blue
      "Blastocyst" = "#984EA3"  # Purple
    ),
    breaks = c("Zygote", "2Cells", "8Cells", "Morula", "Blastocyst"),
    labels = c("Zygote", "2Cell", "8Cell", "Morula", "Blastocyst")
  ) +
  labs(
    x = sprintf("PC1 (%.1f%%)", var_explained[1]),
    y = sprintf("PC2 (%.1f%%)", var_explained[2]),
    color = "Stage"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```

#Doing a PCA for the treatment data plus zygote, eigenvectore generated:
#(aos3) erikson@deep22:/data/jenuwein/erikson/2026/ulATAC/Project3915_treatment_plusZygotes/aos2/bw$ plotPCA --corData results.npz -o PCA.png --colors blue blue blue blue blue blue blue blue blue blue blue blue blue blue blue green green green green green green green green green green green green green green green green orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange orange red red red red red red red red red red red red red red red red red red red red red red red red red red red red red --transpose --ntop 5000 --outFileNameData  PCAeigenvalue_treatmentData_Project3915_plusZygote.tab

```{r}
pca <- read.table("/data/jenuwein/erikson/2026/ulATAC/Project3915_treatment_plusZygotes/aos2/bw/PCAeigenvalue_treatmentData_Project3915_plusZygote.tab", header = TRUE, row.names = 1)
pca_t <- as.data.frame(t(pca))
pca_t <- pca_t[rownames(pca_t) != "Eigenvalue", ]
pca_t$stage <- sub("_rep.*", "", rownames(pca_t))
pca_t$stage <- sub("^X", "", pca_t$stage)
pca_t$stage <- sub("_WT$", "", pca_t$stage)

pca_t$color <- NULL

pca_t$PC1 <- pca_t[,1]
pca_t$PC2 <- pca_t[,2]

pca_t$stage <- gsub("^X", "", pca_t$stage)
pca_t$stage <- factor(pca_t$stage,
                      levels = c("2Cell_ctrl", "2Cell_A485", "2Cells", "Zygote"))

p<-ggplot(pca_t, aes(x = PC1, y = PC2, color = stage)) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "2Cell_ctrl" = "red",
    "2Cell_A485" = "yellow",
    "2Cells" = "green",
    "Zygote" = "purple"
  )) +
  theme_classic()

ggsave("/data/jenuwein/erikson/2026/ulATAC/Project3915_treatment_plusZygotes/PCA.png", plot = p, width = 6, height = 5, dpi = 300)
```

#PCA RNA
```{r}
pca <- read.table("/data/jenuwein/erikson/2026/ulATAC/RNAseq/PCA.bed_coverage.tsv", header = TRUE, row.names = 1)
pca_t <- as.data.frame(t(pca))
pca_t <- pca_t[rownames(pca_t) != "Eigenvalue", ]
#pca_t$stage <- sub("Ctrl_*", "Ctrl", rownames(pca_t))
#pca_t$stage <- sub("Treated_*", "Treated", rownames(pca_t))
pca_t$stage <- sub("_\\d+", "", rownames(pca_t))
pca_t$color <- with(pca_t, ifelse(stage == "Ctrl", "green",
                          ifelse(stage == "Treated", "red", "black")))

library(ggplot2)

pca_t$PC1 <- pca_t[,1]
pca_t$PC2 <- pca_t[,2]
p<-ggplot(pca_t, aes(x = PC1, y = PC2, color = stage)) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "Ctrl" = "green",
    "Treated" = "red"
  )) +
  theme_classic()

```